---
title: "Using OpenTrials API with R"
author: "Darko Bergant"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output: 
  html_document: 
    keep_md: yes
    toc: yes
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# About OpenTrials

From [OpenTrials](http://opentrials.net) page: _OpenTrials is a collaboration between
Open Knowledge and Dr Ben Goldacre from the University of Oxford DataLab. It
aims to locate, match, and share all publicly accessible data and documents, on
all trials conducted, on all medicines and other treatments, globally._

In addition to [web search](http://explorer.opentrials.net/search), the 
OpenTrials database is also accessible via API, documented as Open API (Swagger)
format. 

This short tutorial shows how to use this API from R with 
[rapiclient](https://github.com/bergant/rapiclient).

# Issues

Report **rapiclient** bugs on [rapiclient issues](https://github.com/bergant/rapiclient/issues).
Source R Markdown for this page is [here](https://github.com/bergant/OpenTrialsRDemo).

Report errors with OpenTrials data (for example, if a trial has incorrect
treatments tagged to it) on [this
page](http://explorer.opentrials.net/flag-error).


# Preparation
Install rapiclient from Github:
```{r eval=FALSE}
devtools::install_github("bergant/rapiclient")
```

Use `get_api` and `get_operations` to create R client functions:

```{r read_api, cache=TRUE}
library(rapiclient)
open_trials_api <- get_api("http://api.opentrials.net/v1/swagger.yaml")
open_trials_schemas <- get_schemas(open_trials_api)
open_trials <- get_operations(open_trials_api, handle_response = content_or_stop)

names(open_trials)
```


# Search Trials

`searchTrials` returns trials based on a search query. By default, it’ll search 
in all of a trial’s attributes. Parameter `q` is a 
[ElasticSearch](https://www.elastic.co/guide/en/elasticsearch/reference/2.3/query-dsl-query-string-query.html#query-string-syntax)
query string. Parameters `page` (1 to 100) and `per_page` (10 to 100) are
optional.

```{r cache=TRUE}
trials <- open_trials$searchTrials(q='public_title:(depressive OR depression)')

trials$total_count
length(trials$items)
```

There are `r trials$total_count` trials matching this search query and
the first `r length(trials$items)` already waiting in the `trials$items` list.

Take a look at the richness of the trial search result schema
before using the results:

```{r schema, fig.height=9, fig.width=8}
library(DiagrammeR)
grViz(
  rapiclient::get_schema_graphviz_dot(
    open_trials_api, 
    open_trials_api$definitions$TrialSearchResults
  ) 
)
```

Print some of the trial attributes in a table:

```{r warning=FALSE, message=FALSE}
library(dplyr)
library(knitr)

lapply(trials$items, function(x) { data.frame(
    title = gsub("\n","",x$public_title), 
    source_id = x$source_id,
    sample_size = ifelse(length(x$target_sample_size)==0, NA, x$target_sample_size),
    status = x$status,
    registered = format(as.POSIXct(x$registration_date)),
    publications = length(x$publications),
    stringsAsFactors = FALSE
)}) %>% 
  bind_rows %>% 
  kable(caption = "Table: some data from trials search result")

```

# Tips & Tricks

Print a function to see its documentation. For example: 
```{r}
open_trials$getTrial
```

If parameter name is a reserved word in R, quote it with backticks:
```{r cache=TRUE}
conditions <- 
  open_trials$autocomplete(`in` = "condition", q = "addiction", per_page = 100) %>% 
  getElement("items") %>% 
  bind_rows %>%  
  select(id, name)
```

Note that operations are not vectorised. You have to use the *apply functions to
get the data for arguments with length > 1. For example, count the trials for
each condition:

```{r cache=TRUE}
conditions$trials <- 
  sapply(conditions$id, function(x) {
    open_trials$searchTrials(q = sprintf("conditions.id:(%s)", x))[["total_count"]] 
  })  
  
conditions %>% arrange(desc(trials)) %>% top_n(30, wt = trials) %>% knitr::kable(.)
```

# OpenTrials Operations
From Swagger definition ("http://api.opentrials.net/v1/swagger.yaml") 

```{r results='asis', echo=FALSE}

for(path in open_trials_api$paths) {
  for(action in path) {
    cat("**", action$operationId, "**\n\n", sep = "")
    cat("Description:", action$description, "\n")
    cat("\n")
  }
}

```

# List of Data Sources
OpenTrials data sources:
```{r list_sources, message=FALSE, warning=FALSE, cache=TRUE}
sources <- open_trials$list()
knitr::kable(bind_rows(sources), caption = "Table of sources")
```

# R Packages
**Attached packages:**
```{r list_packages, results='asis', warning=FALSE, cache=TRUE}
sessionInfo()$otherPkgs %>%
  names %>% 
  lapply(citation) %>% 
  lapply(first) %>% 
  lapply(print, style = "html") %>% 
  invisible

```

**Other loaded packages in R session**

```{r results='asis', cache=TRUE}
sessionInfo()$loadedOnly %>%
  lapply(function(x) paste(x$Package, x$Version, x$URL)) %>% 
  paste(collapse = " - ") %>% 
  cat

```



